---
title: "Quick Start Guide for scTypeEval"
author:
  - name: Josep Garnica
    affiliation: University of Geneva
    email: josep.garnicacaparros@unige.ch
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

This quick start guide demonstrates the essential steps for evaluating cell type
annotations using `scTypeEval`. For a comprehensive tutorial, see the main
vignette.


```{r load}
library(scTypeEval)
```

# Minimal Workflow

## From a Count Matrix

```{r minimal_workflow}
library(Matrix)

# Generate example data
set.seed(123)
counts <- Matrix(rpois(50000, 5), nrow=500, ncol=100, sparse=TRUE)
rownames(counts) <- paste0("Gene", seq_len(500))
colnames(counts) <- paste0("Cell", seq_len(100))

metadata <- data.frame(
  celltype = rep(c("TypeA", "TypeB", "TypeC", "TypeD"), each=25),
  sample = rep(paste0("S", seq_len(5)), times=20),
  row.names = colnames(counts)
)

# Create object
sceval <- create_scTypeEval(matrix=counts, metadata=metadata)

# Process data
sceval <- run_processing_data(
  sceval,
  ident = "celltype",
  sample = "sample",
  min_samples = 3,
  min_cells = 5
)

# Identify features
sceval <- run_hvg(sceval,
                  var_method = "basic",
                  ngenes = 1000)

# Run PCA
sceval <- run_pca(sceval, ndim = 20)

# Compute dissimilarity
sceval <- run_dissimilarity(
  sceval,
  method = "Pseudobulk:Euclidean",
  reduction = TRUE
)

# Get consistency
results <- get_consistency(
  sceval,
  dissimilarity_slot = "Pseudobulk:Euclidean",
  consistency_metric = "silhouette"
)
print(results)
```

## From a Seurat Object

```{r seurat_example, eval=FALSE}
library(Seurat)

# Assuming you have a Seurat object named 'seurat_obj'
sceval <- create_scTypeEval(seurat_obj)

# Continue with standard workflow
sceval <- run_processing_data(
  sceval,
  ident = "celltype",      # Adjust to your metadata column
  sample = "orig.ident",   # Adjust to your sample column
  min_samples = 5,
  min_cells = 10
)

sceval <- run_hvg(sceval)
sceval <- run_pca(sceval, ndim = 30)
sceval <- run_dissimilarity(sceval, method = "Pseudobulk:Euclidean")

consistency <- get_consistency(
  sceval,
  dissimilarity_slot = "Pseudobulk:Euclidean",
  consistency_metric = "silhouette"
)
```

## From a SingleCellExperiment Object

```{r sce_example, eval=FALSE}
library(SingleCellExperiment)

# Assuming you have a SingleCellExperiment object named 'sce'
sceval <- create_scTypeEval(sce)

# Continue with workflow as above
```

# Common Use Cases

## Compare Multiple Dissimilarity Methods

```{r compare_methods, eval=FALSE}
# Compute different dissimilarity methods
sceval <- run_dissimilarity(
  sceval,
  method = "Pseudobulk:Euclidean",
  reduction = TRUE
)
sceval <- run_dissimilarity(
  sceval,
  method = "Pseudobulk:Cosine",
  reduction = TRUE
)
sceval <- run_dissimilarity(
  sceval,
  method = "WasserStein",
  reduction = TRUE
)

# Compare consistency across methods
methods <- c("Pseudobulk:Euclidean", "Pseudobulk:Cosine", "WasserStein")
comparison <- lapply(methods, function(m) {
  get_consistency(
    sceval,
    dissimilarity_slot = m,
    consistency_metric = "silhouette"
  )
})
results_df <- do.call(rbind, comparison)
```

## Evaluate Multiple Consistency Metrics

```{r multiple_metrics, eval=FALSE}
# Compute multiple consistency metrics
metrics <- c("silhouette", "NeighborhoodPurity", "Average_similarity")

all_metrics <- lapply(metrics, function(metric) {
  get_consistency(
    sceval,
    dissimilarity_slot = "Pseudobulk:Euclidean",
    consistency_metric = metric
  )
})

combined_results <- do.call(rbind, all_metrics)
```

## Visualize Results

```{r visualize, eval=FALSE}
# Heatmap of dissimilarities
plot_heatmap(
  sceval,
  dissimilarity_slot = "Pseudobulk:Euclidean",
  sort_consistency = "silhouette"
)

plot_pca(
  sceval,
  reduction_slot = "pseudobulk"
)
```

## Using Marker Genes Instead of HVGs

```{r markers, eval=FALSE}
# Identify cell type markers
sceval <- run_gene_markers(
  sceval,
  method = "scran.findMarkers",
  ngenes_celltype = 50
)

# Use markers for dissimilarity calculation
sceval <- run_dissimilarity(
  sceval,
  method = "Pseudobulk:Euclidean",
  gene_list = "Marker_genes",
  reduction = FALSE
)
```

## Focus on Specific Gene Sets

```{r gene_sets, eval=FALSE}
# Add custom gene list
immune_genes <- c("CD3D", "CD4", "CD8A", "CD19", "CD14", "NCAM1")
sceval <- add_gene_list(
  sceval,
  list.name = "immune_markers",
  gene_list = immune_genes
)

# Run analysis on custom genes
sceval <- run_dissimilarity(
  sceval,
  method = "Pseudobulk:Euclidean",
  gene_list = "immune_markers",
  reduction = FALSE
)
```

# Interpreting Results

## What Low Scores Mean

Low consistency scores may indicate:

- **Ambiguous cell type boundaries** between related types
- **Heterogeneous populations** needing refinement
- **Annotation inconsistencies** across samples

## Next Steps for Low-Scoring Cell Types

1. **Visualize** using `plot_heatmap()` or `plot_pca()` to identify
  problematic samples
2. **Investigate** biological differences (e.g., disease vs. healthy)
3. **Refine** annotations by splitting or merging cell types

# Available Methods and Metrics

## Dissimilarity Methods

- **Pseudobulk:Euclidean** - Euclidean distance on pseudobulk profiles
- **Pseudobulk:Cosine** - Cosine distance on pseudobulk profiles
- **Pseudobulk:Pearson** - Pearson correlation distance on pseudobulk profiles
- **WasserStein** - Wasserstein distance between cell distributions
- **recip_classif:Match** - Reciprocal classification matching
- **recip_classif:Score** - Reciprocal classification scoring

## Consistency Metrics

- **silhouette** - Standard silhouette coefficient
- **2label_silhouette** - Two-label silhouette variant
- **NeighborhoodPurity** - K-nearest neighbor purity
- **ward_PropMatch** - Ward clustering proportion match
- **Orbital_medoid** - Medoid-based orbital metric
- **Average_similarity** - Average within-group similarity

# Tips and Best Practices

1. **Always use multiple samples** (minimum 3-5 per cell type)
2. **Compare different methods** - no single method is perfect
3. **Use PCA for speed** - similar results, much faster
4. **Start with HVGs** - then try marker genes if needed
5. **Check sample sizes** - ensure adequate cells per type per sample
6. **Interpret in context** - consider biological heterogeneity

# Getting Help

- **GitHub**: https://github.com/carmonalab/scTypeEval
- **Issues**: https://github.com/carmonalab/scTypeEval/issues
- **Main vignette**: `browseVignettes("scTypeEval")`

# Session Info

```{r session_info}
sessionInfo()
```
